Positivity-Preserving Adaptive Runge--Kutta Methods

### 2020-05-10 nusslein_vsrp/latex/overview.tex
clean up reproducibility notebooks

We say an initial value problem
\begin{align} \label{ode}
    u'(t) & = f(t,u) \\
    u(0) & = u_0
\end{align}
where $u\colon [0,T] \to \R^m$ is positive if
\begin{align} \label{continuous-positivity}
    u(0)\ge 0 \implies u(t) \ge 0 \text{ for all } t \in [0,T].
\end{align}
Here and in the following, inequalities like $u \geq 0$ are meant componentwise.
A sufficient condition for positivity of \eqref{ode} is
\begin{equation} \label{eq:condition_ODE_pos}
u_i=0 \implies f_i(t,u_1,\cdots,u_i,\cdots,u_n) \geq 0 \quad \forall {u \geq 0,}\; \ \  \forall {t\in[0,T]}.
\end{equation}
For such ODEs, the backward Euler method is guaranteed to preserve positivity
under any step size, while the forward Euler will preserve positivity for small enough $\dt$ \cite{hundsdorfer_numerical_2003}.
Any RKM (or in fact any general linear method) that is unconditionally positivity preserving for all positive ODEs
must have order $\le 1$ \cite{bolley_conservation_1978}.
For any higher order method, we expect positivity only under some restriction
on the time step size.

@@

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\State Initialize $n \leftarrow 0$, $u^n \leftarrow u_0$, $t \leftarrow 0$
\While{$t<t_{end}$}
\State Choose $\Delta t$ (fixed or via an adaptive stepsize control)
\State Calculate $F = (f_1, \dots, f_s)$ according to \eqref{eq:stagevalues} \label{line:calc_stge}
\State $u^{n+1} \leftarrow u^n + \Delta t Fb$, cf.\ \eqref{eq:Combination}
\If {$u^{n+1} \geq 0$}
	\State  \textbf{GOTO} line\,\ref{line:update}
\Else
	\State{$\pt \leftarrow p_{start}$}
	\While{$\pt \geq p_{min}$}
		\State Solve LP \eqref{eq:free-adaptation}
		\If {LP is feasible}
			\State $\delta \leftarrow \| \Delta t F(\bt-b)\|$
			\If {$\delta < tol_\delta$}
				\State $u^{n+1} \leftarrow u^n+\Delta t F \tilde b$
				\State  \textbf{GOTO} line\,\ref{line:update}
			\EndIf
		\EndIf
		\State $\pt \leftarrow \pt - 1$
	\EndWhile
	\State Reduce $\dt$
	\State \textbf{GOTO} line\,\ref{line:calc_stge}
\EndIf
\State Estimate $error$ according to \eqref{eq:Err} \label{line:update}
\If {$error \leq tol_{error}$}
	\State $t \leftarrow t + \dt$, $n \leftarrow n+1$
\Else
	\State Reduce $\dt$
	\State \textbf{GOTO} line\,\ref{line:calc_stge}
\EndIf
\EndWhile
\end{algorithmic}
\caption{Pseudocode for the algorithm using a free adaption of weights.}
\label{alg:Adaption}
\end{algorithm}

@@

The total error is split up in the truncation error $err_T$ and the perturbation $\delta$ using the triangle inequality.
The truncation error can be estimated using the standard error estimators $err_T = \| u^{n}_{b} - u^{n}_{\hat{b}} \|$.
After adapting the weights, the perturbation is calculated. If the perturbation is larger than the tolerance, the weights are rejected.
The two values are added to get an approximation of the total error $err = err_T + \delta$.
This type of error estimation is easy to implement because it can be easily
incorporated in an existing step size control and takes advantage of the
standard error approximation.

@@

A very interesting class of methods are the implicit extrapolation methods. These are diagonally implicit and $p < s$. This means that they allow for changes of the weights without a reduction of the order.
Moreover, all stage values are computed using the BE method. Hence, all intermediate stages are positive.
Furthermore, an embedded BE step is included. This ensures that a positive solution always exists, even if it is of first order.

@@

\begin{figure}
\centering
\includegraphics[width=0.85\textwidth]{plots/ADP_stepsize,b.pdf}
\caption{Statistics of computation of the advection-diffusion-reaction problem \eqref{eq:ADR}.}
\label{fig:Stats_ADP}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.85\textwidth]{plots/Stratospheric_stepsize,b.pdf}
\caption{Statistics of computation of the stratospheric reaction system
         \eqref{eq:stratospheric}.}
\label{fig:Stats_Strat}
\end{figure}

>>>>

We say an initial value problem
\begin{align} \label{ode}
    u'(t) & = f(t,u) \\
    u(0) & = u_0
\end{align}
where $u\colon [0,T] \to \R^m$ is positive if
\begin{align} \label{continuous-positivity}
    u(0)\ge 0 \implies u(t) \ge 0 \text{ for all } t \in [0,T].
\end{align}
Here and in the following, inequalities like $u \geq 0$ are meant componentwise.
A sufficient condition for positivity of \eqref{ode} is
\begin{equation} \label{eq:condition_ODE_pos}
u_i=0 \implies f_i(t,[u_1,\cdots,u_i,\cdots,u_n]^T) \geq 0 \quad \forall {u \geq 0,}\; \ \  \forall {t\in[0,T]}.
\end{equation}
For such ODEs, the backward Euler method is guaranteed to preserve positivity
under any step size, while the forward Euler will preserve positivity for small enough $\dt$ \cite{hundsdorfer_numerical_2003}.
Any RKM (or in fact any general linear method) that is unconditionally positivity preserving for all positive ODEs
must have order $\le 1$ \cite{bolley_conservation_1978}.
For any higher order method, we expect positivity only under some restriction
on the time step size.

@@

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\State Initialize $n \leftarrow 0$, $u^n \leftarrow u_0$, $t \leftarrow 0$
\While{$t<t_{end}$}
\State Choose $\Delta t$ (fixed or via an adaptive stepsize control)
\State Calculate $F = (f_1, \dots, f_s)$ according to \eqref{eq:stagevalues} \label{line:calc_stge}
\State $u^{n+1} \leftarrow u^n + \Delta t Fb$, according to \eqref{eq:Combination}
\If {$u^{n+1} \geq 0$}
	\State  \textbf{GOTO} line\,\ref{line:update}
\Else
	\State{$\pt \leftarrow p_{start}$}
	\While{$\pt \geq p_{min}$}
		\State Solve LP \eqref{eq:free-adaptation}
		\If {LP is feasible}
			\State $\delta \leftarrow \| \Delta t F(\bt-b)\|$
			\If {$\delta < tol_\delta$}
				\State $u^{n+1} \leftarrow u^n+\Delta t F \tilde b$
				\State  \textbf{GOTO} line\,\ref{line:update}
			\EndIf
		\EndIf
		\State $\pt \leftarrow \pt - 1$
	\EndWhile
	\State Reduce $\dt$
	\State \textbf{GOTO} line\,\ref{line:calc_stge}
\EndIf
\State Estimate $error$ according to \eqref{eq:Err} \label{line:update}
\If {$error \leq tol_{error}$}
	\State $t \leftarrow t + \dt$, $n \leftarrow n+1$
\Else
	\State Reduce $\dt$
	\State \textbf{GOTO} line\,\ref{line:calc_stge}
\EndIf
\EndWhile
\end{algorithmic}
\caption{Pseudocode for the algorithm using a free adaption of weights.}
\label{alg:Adaption}
\end{algorithm}

@@

The total error is split up in the truncation error and the perturbation $\delta$ using the triangle inequality.
The truncation error can be estimated using the standard error estimators $err_T = \| u^{n}_{b} - u^{n}_{\hat{b}} \|$.
After adapting the weights, the perturbation is calculated. If the perturbation is larger than the tolerance, the weights are rejected.
The two values are added to get an approximation of the total error $err = err_T + \delta$.
This type of error estimation is easy to implement because it can be easily
incorporated in an existing step size control and takes advantage of the
standard error approximation.

@@

A very interesting class of methods are the implicit extrapolation methods. These allow changes of the weights without a reduction of the order, as discussed in Section\,\ref{sec:integration}.
Moreover, all stage values are computed using the BE method. Hence, all intermediate stages are positive.
Furthermore, an embedded BE step is included. This ensures that a positive solution always exists, even if it is of first order.

@@

\begin{figure}
\centering
\includegraphics[width=0.85\textwidth]{plots/ADP_stepsize_b.pdf}
\caption{Statistics of computation of the advection-diffusion-reaction problem \eqref{eq:ADR}.}
\label{fig:Stats_ADP}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.85\textwidth]{plots/Stratospheric_stepsize_b.pdf}
\caption{Statistics of computation of the stratospheric reaction system
         \eqref{eq:stratospheric}.}
\label{fig:Stats_Strat}
\end{figure}
